A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-15 tidyindex.qmd
clean up after working with Patricia

In what follows when geographical or temporal components of the $x(s,t)$ process are fixed we will be using suffix notation. For example, $x_{s}(t)$ represents the data for a fixed location $s$ as a function of time $t$. While $x_{t}(s)$ denotes the spatial varying procces for a fixed $t$.

<!-- We assume the spatial and temporal indexes do not depend on the variable, that is, all the variables share the same set of spatial and temporal indexes.-->

<!--For example, a year-month combination, $(Y, M)^\prime$, can be used to record monthly precipitation series from January 1990 to December 2020, where $Y = 1990, \cdots, 2020$ and $M = 1, \cdots, 12$. .-->

<!-- To simplify the notation, a subscript letter $s$ or $t$ is used to denote the data have a fixed spatial locations or temporal order. When the spatial locations are fixed, the data become multivariate time series, denoted as $x^l_s(t)$. Similarly,  $x^l_t(s)$ represents multivariate spatial data when the temporal dimension is fixed. -->

| No. | Module                  | Input                                                                  | Operation                                                                                                                                          | Output                                                                 |
|-------------|-------------|-------------|--------------------|-------------|
| 1   | Temporal processing     | $x_{s}(t)$                                                             | $f_{\Psi}[x(.)]$                                                                                                                                   | $\tilde{x}^i(t^\prime)$ where $t\prime \in D_{t^\prime}$                       |
| 2   | Spatial processing      | $x_t(s)$                                                               | $g[x(.)]$                                                                                                                                 | $x^i(s^\prime)$ where $s^\prime \in D_{s^\prime}$                     |
| 3   | Variable transformation | $x(s; t)$                                                              | $T(x(s;t))$                                                                                                                                        | $Tx^i_t(s)$ ????                                                          |
| 4   | Scaling                 | $x(s; t)$                                                              | $[x(s;t) - \alpha]/\gamma$                                                                                                                         | $Sx^i(s;t)$ where bbbbb                                                   |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $\mathbf{y} \subseteq \mathbb{R}^d$ and$d < q$ |
| 6   | Distribution fit        | $x(s; t)$                                                              | $F_\mathbf{\theta}(x(s,t))$ where $F_{\mathbf{\theta}}(.)$ is the cumulated distribution function with distribution parameter(s) $\mathbf{\theta}$ | $p(s,t)$ where $p(.)$ denotes the probability value between 0 and 1    |
| 7   | Normalising             | $x(s; t)$                                                              | $\Phi^{-1}[x(s, t)]$                                                                                                                               | $z(s; t)$ where $z \sim N(.,.)$                                        |
| 8   | Benchmarking            | $x(s; t)$                                                              |                                                                                                                                                    | $b(s;t)$                                                               |
| 9   | Simplification          | $x(s; t)$                                                              | ccccc                                                                                                                                              | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |

: An notation overview of the input, operation, and output of each pipeline module.

Some notes:

-   still $x(s,t)$ notation because for all the other modules, $s$ and $t$ are all fixed, then $x(s, t)$ becomes $x$????
-   aaaaa: just a general function over the input????
-   bbbbb: input and output has the same distributional shape
-   ccccc: latex issue

```{r fig-pipeline-steps}
#| fig-align: center
#| fig-cap: "Diagram of pipeline steps for index construction. will need to be updated with better design and the distribution fitting step."
#| out-height: 90%
#| out-width: 100%
knitr::include_graphics(here::here("figures/pipeline-overall.png"))
```

@@

`Input`: $x(t)$, time series of a single variable with fixed locations

`Output`: $x^*(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.

`Computation`: The temporal processing module takes as input argument a single variable $x_s^i(t)$ at location $s$  as a function of time. In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$,  $\tilde{x}_s^i(t^') = f[[x_s(t)]$ where $t^'$ refers to the new temporal resolution after aggregation. An example of a temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$. That is also known as the time scale. 
<!-- Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated precipitation series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be expressed as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.  -->
For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).

<!-- to incorporate the temporal neighbouring effect: $x^*(.) = f_{\Psi}[x(.)]$, where $\Psi$ is a vector of hyper-parameters associated with the transformation. In temporal processing, the transition of temporal dimension from $t$ to $t^\prime \subseteq D_{t^\prime}$ allows for the decomposition of time into its constituent temporal components. For example,  -->

<!-- The construction of an index sometimes needs to consider information from neighbouring time periods. The temporal processing is a general operator on the time dimension of the data in the form of -->

<!-- ```{=tex} -->

<!-- x(\mathbf{s};\mathbf{t}^\prime) = f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})), -->

<!-- ``` -->

<!-- where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. -->

@@

`Input`: $x(s)$

`Output`: $x^\#(s^\prime)$ where $s^\prime \in D_{s^{\prime}}$

`Computation`: The spatial processing module takes a single variable with a fixed temporal dimension, $x(s)$, as input. This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^\#(.) = g_{\Theta}(x(.))$, where $\Theta$ is a vector of hyper-parameters associated with the transformation. The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions. This also includes the aggregation of variables into different levels, such as city, state, and countries scales.

<!-- may be needed when indexes are not calculated independently on each collected location or when variables collected from multiple sources need to be fused before further processing. The process can be written as a general operation in the form of -->

<!-- ```{=tex} -->

<!-- x(\mathbf{s}^\prime;\mathbf{t}) = g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})), -->

<!-- ``` -->

<!-- where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the associated parameters in the process and $d_{\theta}$ is the number of parameter of $\theta$. An example of spatial processing is to align variables collected in different resolutions. When variables are collected at different resolutions, analysts may choose to down-sample those in a finer resolution, $i$, to match those in a coarser resolution, $i^\prime$. This is a spatial aggregation and if aggregate using the mean, it can be written as -->

<!-- ```{=tex} -->

<!-- g(x) = \frac{\sum_{i \in i^\prime}x}{n_{i^\prime}}, -->

<!-- ``` -->

<!-- where $i \in i^\prime$ includes all the cells from the finer resolution in the coarser grid and $n_{i^\prime}$ is the number of observations falls into the coarser grid. Other examples of spatial processing include 1) borrowing information from neighbouring spatial locations to interpolate unobserved locations and 2) fusing variables from ground measures with satellite imageries. -->

@@

`Input`: $\mathbf{x}(s;t)$ where $\mathbf{x} \subseteq \mathbb{R}^p$

`Output`: $\mathbf{y}(s;t)$ where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < p$

`Computation`: Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$ , as the input. It summarises the high-dimensional information into a lower dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < p$, as the output.

The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided in statistical methods. For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$). This is the only step that differs from the Standardized Precipitation Index (SPI).

(Move this to the example and remove notation)
When a physical process is absent to describe the phenomenon, multivariate data are often combined using statistical methods, which can be further classified into linear and non-linear techniques. Many social indexes use a linear combination in an additive structure, $y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),$ where $\lambda_i$ is the weight assigned to each individual variable $x_i$. A linear combination can also be preceived as a linear projection of multivariate information through the weight vector. Projecting data from higher to lower dimension unavoidably leads to information loss and the used weights need careful examination to understand their effects on the index and implications for interpretation and decision-making. One can visualise the data projection across different sets of weights through animation using a method called tour.

@@

@@

```         
gggi %>% 
  init(id = country) %>%
  add_meta(gggi_weights, var_col = variable) %>% 
  dimension_reduction(
    index_new = aggregate_linear(
      ~labour_force_participation:years_with_female_head_of_state,
      weight = weight)) 
```

>>>>

In what follows when geographic or temporal components of the $x(s,t)$ process are fixed we will be using suffix notation. For example, $x_{s}(t)$ represents the data for a fixed location $s$ as a function of time $t$. While $x_{t}(s)$ denotes the spatial varying proccess for a fixed $t$. An overview of the notation for pipeline input, operation, and output is present in @tbl-notation.

| No. | Module                  | Input                                                                  | Operation                                                                         | Output                                                                 |
|-|-------------------|--------------|----------------------|----------------------|
| 1   | Temporal processing     | $x^i_{s}(t)$                                                           | $f[x^i_s(t)]$                                                                         | $\tilde{x}^i_s(t^\prime)$, where $t^\prime \in D_{t^\prime}$            |
| 2   | Spatial processing      | $x^i_t(s)$                                                             | $g[x^i_t(s)]$                                                                         | $x^i_t(s^\prime)$, where $s^\prime \in D_{s^\prime}$ ????                   |
| 3   | Variable transformation | $x^i(s; t)$                                                            | $T[x(s;t)]$                                                                       | $Tx^i_t(s)$ ????                                                       |
| 4   | Scaling                 | $x^i(s; t)$                                                            | $[x(s;t) - \alpha]/\gamma$                                                        | $Sx^i(s;t)$ ????                                                       |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                                                              | $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < q$ |
| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x(s;t)]$, where $F_{\mathbf{\theta}}(.)$ is the CDF function | $p(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
| 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x(s; t)]$                                                              | $z(s; t)$, where $z$ has a normal distribution                          |
| 8   | Benchmarking            | $x^i(s; t)$                                                            | $u[x(s;t)]$                                                                       | $b(s;t)$                                                               |
| 9   | Simplification          | $x^i(s; t)$                                                            | ccccc                                                                             | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |

: An notation overview of the input, operation, and output of each pipeline module. {#tbl-notation}

```{r fig-pipeline-steps}
#| fig-align: center
#| fig-cap: "Diagram of pipeline modules for index construction. The highlighted path illustrates one possible construction using the dimension reduction and simplification module."
#| out-height: 90%
#| out-width: 100%
knitr::include_graphics(here::here("figures/pipeline-overall.png"))
```

@@

The temporal processing module takes as input argument a single variable $x_s^i(t)$ at location $s$ as a function of time. In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$, $\tilde{x}_s^i(t^\prime) = f[x^i_s(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation. An example of a temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$. That is also known as the time scale. For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).

<!-- Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated precipitation series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be expressed as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.  -->

@@

The spatial processing module takes a single variable with a fixed temporal dimension, $x(s)$, as input. This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^\#(.) = g_{\Theta}(x(.))$, where $\Theta$ is a vector of hyper-parameters associated with the transformation. The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions. This also includes the aggregation of variables into different levels, such as city, state, and countries scales.

@@

Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$ , as the input. It summarises the high-dimensional information into a lower dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < p$, as the output. The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided in statistical methods. For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$). This is the only step that differs from the Standardized Precipitation Index (SPI).

@@

(Moved from the pipeline module to here) When a physical process is absent to describe the phenomenon, multivariate data are often combined using statistical methods, which can be further classified into linear and non-linear techniques. Many social indexes use a linear combination in an additive structure, $y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),$ where $\lambda_i$ is the weight assigned to each individual variable $x_i$. A linear combination can also be preceived as a linear projection of multivariate information through the weight vector. Projecting data from higher to lower dimension unavoidably leads to information loss and the used weights need careful examination to understand their effects on the index and implications for interpretation and decision-making. One can visualise the data projection across different sets of weights through animation using a method called tour.

@@

    gggi %>% 
      init(id = country) %>%
      add_meta(gggi_weights, var_col = variable) %>% 
      dimension_reduction(
        index_new = aggregate_linear(
          ~labour_force_participation:years_with_female_head_of_state,
          weight = weight))

### 2023-08-15 tidyindex.tex
clean up after working with Patricia

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, sharp corners, enhanced, frame hidden]}{\end{tcolorbox}}\fi

@@

In what follows when geographical or temporal components of the
\(x(s,t)\) process are fixed we will be using suffix notation. For
example, \(x_{s}(t)\) represents the data for a fixed location \(s\) as
a function of time \(t\). While \(x_{t}(s)\) denotes the spatial varying
procces for a fixed \(t\).

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2778}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1806}}@{}}
\caption{An notation overview of the input, operation, and output of
each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x_{s}(t)\) & \(f_{\Psi}[x(.)]\) &
\(\tilde{x}^i(t^\prime)\) where \(t\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x_t(s)\) & \(g[x(.)]\) & \(x^i(s^\prime)\)
where \(s^\prime \in D_{s^\prime}\) \\
3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) &
\(Tx^i_t(s)\) ???? \\
4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(Sx^i(s;t)\)
where bbbbb \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\) where
\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) & aaaaa &
\(\mathbf{y}(s;t)\) where \(\mathbf{y} \subseteq \mathbb{R}^d\)
and\(d < q\) \\
6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
\(F_{\mathbf{\theta}}(.)\) is the cumulated distribution function with
distribution parameter(s) \(\mathbf{\theta}\) & \(p(s,t)\) where
\(p(.)\) denotes the probability value between 0 and 1 \\
7 & Normalising & \(x(s; t)\) & \(\Phi^{-1}[x(s, t)]\) & \(z(s; t)\)
where \(z \sim N(.,.)\) \\
8 & Benchmarking & \(x(s; t)\) & & \(b(s;t)\) \\
9 & Simplification & \(x(s; t)\) & ccccc &
\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

Some notes:

\begin{itemize}
\tightlist
\item
  still \(x(s,t)\) notation because for all the other modules, \(s\) and
  \(t\) are all fixed, then \(x(s, t)\) becomes \(x\)????
\item
  aaaaa: just a general function over the input????
\item
  bbbbb: input and output has the same distributional shape
\item
  ccccc: latex issue
\end{itemize}

@@

\caption{\label{fig-pipeline-steps}Diagram of pipeline steps for index
construction. will need to be updated with better design and the
distribution fitting step.}

@@

\texttt{Input}: \(x(t)\), time series of a single variable with fixed
locations

\texttt{Output}: \(x^*(t^{\prime})\), where
\(t^\prime \in D_{t^{\prime}}\).

\texttt{Computation}: The temporal processing module takes as input
argument a single variable \(x_s^i(t)\) at location \(s\) as a function
of time. In this step the original time series can be transformed or
summarized into a new one via time aggregation, the transformation is
represented by the function \(f\),
\(\tilde{x}_s^i(t^\prime) = f[[x_s(t)]\) where \(t^\prime\) refers to
the new temporal resolution after aggregation. An example of a temporal
processing done in the computation of the Standardized Precipitation
Index (SPI), consists of summing the monthly precipitation series over a
rolling time window of size \(k\). That is also known as the time scale.
For SPI, the choice of the time scale \(k\) is used to control the
accumulation period for the water deficit, enabling the assessment of
drought severity across various types (meteorological, agricultural, and
hydrological).

@@

\texttt{Input}: \(x(s)\)

\texttt{Output}: \(x^\#(s^\prime)\) where
\(s^\prime \in D_{s^{\prime}}\)

\texttt{Computation}: The spatial processing module takes a single
variable with a fixed temporal dimension, \(x(s)\), as input. This step
transforms the variable from the original spatial dimension \(s\) into
the new dimension \(s^\prime \in D_{s^\prime}\) through
\(x^\#(.) = g_{\Theta}(x(.))\), where \(\Theta\) is a vector of
hyper-parameters associated with the transformation. The change of
spatial dimension allows for the alignment of variables collected from
different measurements, such as in-situ stations and satellite imagery,
or originating from different resolutions. This also includes the
aggregation of variables into different levels, such as city, state, and
countries scales.

@@

\texttt{Input}: \(\mathbf{x}(s;t)\) where
\(\mathbf{x} \subseteq \mathbb{R}^p\)

\texttt{Output}: \(\mathbf{y}(s;t)\) where
\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < p\)

\texttt{Computation}: Dimension reduction takes the multivariate
information \(\mathbf{x}(s;t)\), where
\(\mathbf{x} \subseteq \mathbb{R}^p\) , as the input. It summarises the
high-dimensional information into a lower dimension representation
\(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\) and \(d < p\),
as the output.

The transformation can be based on domain-specific knowledge,
originating from theories describing the underlying physical processes,
or guided in statistical methods. For example, the Standardized
Precipitation-Evapotranspiration Index (SPEI) calculates the difference
\(D\) between precipitation (\(P\)) and potential evapotranspiration
(\(\text{PET}\)), using a water balance model (\(D = P - \text{PET}\)).
This is the only step that differs from the Standardized Precipitation
Index (SPI).

(Move this to the example and remove notation) When a physical process
is absent to describe the phenomenon, multivariate data are often
combined using statistical methods, which can be further classified into
linear and non-linear techniques. Many social indexes use a linear
combination in an additive structure,
\(y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),\) where \(\lambda_i\)
is the weight assigned to each individual variable \(x_i\). A linear
combination can also be preceived as a linear projection of multivariate
information through the weight vector. Projecting data from higher to
lower dimension unavoidably leads to information loss and the used
weights need careful examination to understand their effects on the
index and implications for interpretation and decision-making. One can
visualise the data projection across different sets of weights through
animation using a method called tour.

@@

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, sharp corners, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

In what follows when geographic or temporal components of the \(x(s,t)\)
process are fixed we will be using suffix notation. For example,
\(x_{s}(t)\) represents the data for a fixed location \(s\) as a
function of time \(t\). While \(x_{t}(s)\) denotes the spatial varying
proccess for a fixed \(t\). An overview of the notation for pipeline
input, operation, and output is present in Table~\ref{tbl-notation}.

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0128}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1795}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x^i_{s}(t)\) & \(f[x^i_s(t)]\) &
\(\tilde{x}^i_s(t^\prime)\), where \(t^\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x^i_t(s)\) & \(g[x^i_t(s)]\) &
\(x^i_t(s^\prime)\), where \(s^\prime \in D_{s^\prime}\) ???? \\
3 & Variable transformation & \(x^i(s; t)\) & \(T[x(s;t)]\) &
\(Tx^i_t(s)\) ???? \\
4 & Scaling & \(x^i(s; t)\) & \([x(s;t) - \alpha]/\gamma\) &
\(Sx^i(s;t)\) ???? \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\), where
\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) &
\(h[\mathbf{x}(s;t)]\) & \(\mathbf{y}(s;t)\), where
\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < q\) \\
6 & Distribution fit & \(x^i(s; t)\) & \(F[x(s;t)]\), where
\(F_{\mathbf{\theta}}(.)\) is the CDF function & \(p(s;t)\), where
\(p(.) \in [0, 1]\) is a probability value \\
7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x(s; t)]\) & \(z(s; t)\),
where \(z\) has a normal distribution \\
8 & Benchmarking & \(x^i(s; t)\) & \(u[x(s;t)]\) & \(b(s;t)\) \\
9 & Simplification & \(x^i(s; t)\) & ccccc &
\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

\caption{\label{fig-pipeline-steps}Diagram of pipeline modules for index
construction. The highlighted path illustrates one possible construction
using the dimension reduction and simplification module.}

@@

The temporal processing module takes as input argument a single variable
\(x_s^i(t)\) at location \(s\) as a function of time. In this step the
original time series can be transformed or summarized into a new one via
time aggregation, the transformation is represented by the function
\(f\), \(\tilde{x}_s^i(t^\prime) = f[x^i_s(t)]\) where \(t^\prime\)
refers to the new temporal resolution after aggregation. An example of a
temporal processing done in the computation of the Standardized
Precipitation Index (SPI), consists of summing the monthly precipitation
series over a rolling time window of size \(k\). That is also known as
the time scale. For SPI, the choice of the time scale \(k\) is used to
control the accumulation period for the water deficit, enabling the
assessment of drought severity across various types (meteorological,
agricultural, and hydrological).

@@

The spatial processing module takes a single variable with a fixed
temporal dimension, \(x(s)\), as input. This step transforms the
variable from the original spatial dimension \(s\) into the new
dimension \(s^\prime \in D_{s^\prime}\) through
\(x^\#(.) = g_{\Theta}(x(.))\), where \(\Theta\) is a vector of
hyper-parameters associated with the transformation. The change of
spatial dimension allows for the alignment of variables collected from
different measurements, such as in-situ stations and satellite imagery,
or originating from different resolutions. This also includes the
aggregation of variables into different levels, such as city, state, and
countries scales.

@@

Dimension reduction takes the multivariate information
\(\mathbf{x}(s;t)\), where \(\mathbf{x} \subseteq \mathbb{R}^p\) , as
the input. It summarises the high-dimensional information into a lower
dimension representation \(\mathbf{y}(s;t)\), where
\(y \subseteq \mathbb{R}^d\) and \(d < p\), as the output. The
transformation can be based on domain-specific knowledge, originating
from theories describing the underlying physical processes, or guided in
statistical methods. For example, the Standardized
Precipitation-Evapotranspiration Index (SPEI) calculates the difference
\(D\) between precipitation (\(P\)) and potential evapotranspiration
(\(\text{PET}\)), using a water balance model (\(D = P - \text{PET}\)).
This is the only step that differs from the Standardized Precipitation
Index (SPI).

@@

(Moved from the pipeline module to here) When a physical process is
absent to describe the phenomenon, multivariate data are often combined
using statistical methods, which can be further classified into linear
and non-linear techniques. Many social indexes use a linear combination
in an additive structure,
\(y(s; t) = \sum_{i = 1}^{p}\lambda_{i}x_i(s;t),\) where \(\lambda_i\)
is the weight assigned to each individual variable \(x_i\). A linear
combination can also be preceived as a linear projection of multivariate
information through the weight vector. Projecting data from higher to
lower dimension unavoidably leads to information loss and the used
weights need careful examination to understand their effects on the
index and implications for interpretation and decision-making. One can
visualise the data projection across different sets of weights through
animation using a method called tour.
